---
title: "STRchive Loci: HPO Terms Data Analysis"
format: html
editor: visual
---

## Importing STRchive-loci.json Data and Accessing HPO Terms

Going to read STRchive data from GitHub into R using the `fromJSON()` function from the `jsonlite` package. Then will view the data using the `View()` function and access the nested HPO terms from the data using the `$` command.

```{r}
#| label: read-jsonlite

# Load packages
library(jsonlite)

# Read data from url into an object called "loci"
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# View the data
View(loci)

# Select HPO terms
loci$hpo_terms
```

## Example: Creating a Plot with STR Motif Length Data

Create a bar plot using the `table()` and `barplot()` functions.

```{r}
#| label: create bar plot

# Load packages
library(jsonlite)

#Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Create barplot
counts <- table(loci$motif_len)
barplot(counts, main = "STRchive loci by motif length",
        xlab = "Motif length (bp)",
        ylab = "Number of loci",
        col = "pink")
```

## Consolidating the "loci" Data Frame into HPO Terms

Going to remove columns from "loci" data using the `select()` function from the `dplyr` package so that only disease_id and hpo_terms are included.

```{r}
#| label: consolidate "loci" data into hpo terms

# Load packages
library(dplyr)
library(jsonlite)

# Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Remove columns from loci data table
hpo_terms <- loci %>% select(c(disease_id, hpo_terms))
View(hpo_terms)
```

## Unnesting and sorting "hpo_terms" from hpo_terms data frame

Going to unnest the hpo_terms from the disease_id rows in the hpo_terms data frame using the `unnest_longer()` function from the `tidyverse` package so that the individual phenotypes for each disease are accessible. Also going to use `relocate` to rearrange data frame with hpo_terms at the front. Finally, going to create a separate data frame of phenotypes sorted by frequency using the `count()` function.

```{r}
#| label: unnest "hpo_terms" values from "disease_id" rows

# Load packages
library(jsonlite)
library(tidyverse)

# Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Remove columns from loci data table and create new hpo_terms data frame
hpo_terms <- loci %>% select(c(disease_id, hpo_terms))
View(hpo_terms)

# Unnest hpo_terms, relocate hpo_terms, and create new unnested_hpo_terms data frame
unnested_hpo_terms <- hpo_terms |> unnest_longer(hpo_terms) |> relocate(hpo_terms)
view(unnested_hpo_terms)

# Sort hpo_terms by frequency
sorted_hpo_terms <- unnested_hpo_terms |> count(hpo_terms, sort = T)
view(sorted_hpo_terms)
```

## Visualizing hpo_terms

Different visualizations of hpo_terms using different `ggplot` functions form the `gglot2` package.

```{r}
#| label: visualize "hpo_terms" as different plots

# Load packages
library(jsonlite)
library(tidyverse)

# Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Consolidate and unnest hpo_terms
hpo_terms <- loci |> select(c(disease_id, hpo_terms))
unnested_hpo_terms <- hpo_terms |> unnest_longer(hpo_terms) |> relocate(hpo_terms)

# Visualize all hpo_terms vs. frequency as a bar plot
unnested_hpo_terms |> ggplot(aes(x = hpo_terms)) + geom_bar(colour = "lightgreen") + labs(title = "HPO Frequency", subtitle = "Frequency of phenotypes across different STR diseases", x = "HPO Term", y = "Frequency")

# Visualize hpo_terms with a frequency of 13+ across diseases
most_frequent_hpo <- unnested_hpo_terms |> count(hpo_terms, sort = T) |> filter(n >= 13)
ggplot(most_frequent_hpo, aes(x = n, y = fct_infreq(hpo_terms, (n)) |> fct_rev())) + geom_col(fill = "lightblue") + labs(title = "Most Frequent HPOs", subtitle = "Phenotypes with a frequency of 13+ across STR diseases", x = "Frequency", y = "HPO Term")

## FINAL FREQUENCY PLOT
# Visualize hpo_terms with a frequency of 13+ across diseases and with "Ataxia" types consolidated (Based on "children" using `ontologyIndex` package with `data(hpo)`
hpo_frequency <- unnested_hpo_terms |> mutate(
  hpo_terms = fct_recode(hpo_terms,
  "HP:0001251 Ataxia" = "HP:0001251 Ataxia",
  "HP:0001251 Ataxia" = "HP:0001310 Dysmetria",
  "HP:0001251 Ataxia" = "HP:0002066 Gait ataxia",
  "HP:0001251 Ataxia" = "HP:0002070 Limb ataxia",
  "HP:0001251 Ataxia" = "HP:0002073 Progressive cerebellar ataxia",
  "HP:0001251 Ataxia" = "HP:0002075 Dysdiadochokinesis",
  "HP:0001251 Ataxia" = "HP:0002078 Truncal ataxia",
  "HP:0001251 Ataxia" = "HP:0010867 Dyssynergia",
  "HP:0001251 Ataxia" = "HP:0002070 Limb Ataxia",
  )) |> count(hpo_terms, sort = T) |> filter (n >= 13)
ggplot(hpo_frequency, aes(x = n, y = fct_infreq(hpo_terms, (n)) |> fct_rev())) + geom_col(fill = "lightblue") + labs(title = "Most Frequent HPOs", subtitle = "Phenotypes with a frequency of 13+ across STR diseases", x = "Frequency", y = "HPO Term")

```

## Visualizing STR motifs (incomplete)

Visualizations of pathogenic_motif_reference_orientation using `ggplot2` functions

```{r}
#| label: visualize "pathogenic_motif_reference_orientation" motifs


# Load packages
library(jsonlite)
library(tidyverse)

# Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Select relevant data for heat map visualization
hpo_hm_data <- loci |> select(hpo_terms, disease_id, pathogenic_motif_reference_orientation)
view(hpo_hm_data)

# Select the most frequent motifs in "pathogenic_motif_reference_orientation" data column (based on frequency across disease_id)
most_frequent_motifs <- loci |> select(pathogenic_motif_reference_orientation, disease_id) |> count(pathogenic_motif_reference_orientation, sort = T)
view(most_frequent_motifs)
```