---
title: "STRchive Loci: HPO Terms Data Analysis"
format: html
editor: visual
---

## Importing STRchive-loci.json Data and Accessing HPO Terms

Going to read STRchive data from GitHub into R using the `fromJSON()` function from the `jsonlite` package. Then will view the data using the `View()` function and access the nested HPO terms from the data using the `$` command.

```{r}
#| label: read-jsonlite

# Load packages
library(jsonlite)

# Read data from url into an object called "loci"
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# View the data
View(loci)

# Select HPO terms
loci$hpo_terms
```

## Example: Creating a Plot with STR Motif Length Data

Create a bar plot using the `table()` and `barplot()` functions.

```{r}
#| label: create bar plot

# Load packages
library(jsonlite)

#Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Create barplot
counts <- table(loci$motif_len)
barplot(counts, main = "STRchive loci by motif length",
        xlab = "Motif length (bp)",
        ylab = "Number of loci",
        col = "pink")
```

## Consolidating the "loci" Data Frame into HPO Terms

Going to remove columns from "loci" data using the `select()` function from the `dplyr` package so that only disease_id and hpo_terms are included.

```{r}
#| label: consolidate "loci" data into hpo terms

# Load packages
library(dplyr)
library(jsonlite)

# Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Remove columns from loci data table
hpo_terms <- loci %>% select(c(disease_id, hpo_terms))
View(hpo_terms)
```

## Unnesting and sorting "hpo_terms" from hpo_terms data frame

Going to unnest the hpo_terms from the disease_id rows in the hpo_terms data frame using the `unnest_longer()` function from the `tidyverse` package so that the individual phenotypes for each disease are accessible. Also going to use `relocate` to rearrange data frame with hpo_terms at the front. Finally, going to create a separate data frame of phenotypes sorted by frequency using the `count()` function.

```{r}
#| label: unnest "hpo_terms" values from "disease_id" rows

# Load packages
library(jsonlite)
library(tidyverse)

# Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Remove columns from loci data table and create new hpo_terms data frame
hpo_terms <- loci %>% select(c(disease_id, hpo_terms))
View(hpo_terms)

# Unnest hpo_terms, relocate hpo_terms, and create new unnested_hpo_terms data frame
unnested_hpo_terms <- hpo_terms |> unnest_longer(hpo_terms) |> relocate(hpo_terms)
view(unnested_hpo_terms)

# Sort hpo_terms by frequency
sorted_hpo_terms <- unnested_hpo_terms |> count(hpo_terms, sort = T)
view(sorted_hpo_terms)
```

## Visualizing hpo_terms

Different visualizations of hpo_terms using different `ggplot` functions form the `gglot2` package.

```{r}
#| label: visualize "hpo_terms" as different plots

# Load packages
library(jsonlite)
library(tidyverse)
library(ontologyIndex)

# Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Consolidate and unnest hpo_terms
hpo_terms <- loci |> select(c(disease_id, hpo_terms))
unnested_hpo_terms <- hpo_terms |> unnest_longer(hpo_terms) |> relocate(hpo_terms)

# Visualize all hpo_terms vs. frequency as a bar plot
unnested_hpo_terms |> ggplot(aes(x = hpo_terms)) + geom_bar(colour = "lightgreen") + labs(title = "HPO Frequency", subtitle = "Frequency of phenotypes across different STR diseases", x = "HPO Term", y = "Frequency")

# Visualize hpo_terms with a frequency of 13+ across diseases
most_frequent_hpo <- unnested_hpo_terms |> count(hpo_terms, sort = T) |> filter(n >= 13)
ggplot(most_frequent_hpo, aes(x = n, y = fct_infreq(hpo_terms, (n)) |> fct_rev())) + geom_col(fill = "lightblue") + labs(title = "Most Frequent HPOs", subtitle = "Phenotypes with a frequency of 13+ across STR diseases", x = "Frequency", y = "HPO Term")

# Visualize hpo_terms with a frequency of 13+ across diseases and with "Ataxia" types consolidated (Based on "children", or `get_descendants()` using `ontologyIndex` package with `data(hpo)`
hpo_frequency <- unnested_hpo_terms |> mutate(
  hpo_terms = fct_recode(hpo_terms,
  "HP:0001251 Ataxia" = "HP:0001251 Ataxia",
  "HP:0001251 Ataxia" = "HP:0001310 Dysmetria",
  "HP:0001251 Ataxia" = "HP:0002066 Gait ataxia",
  "HP:0001251 Ataxia" = "HP:0002070 Limb ataxia",
  "HP:0001251 Ataxia" = "HP:0002073 Progressive cerebellar ataxia",
  "HP:0001251 Ataxia" = "HP:0002075 Dysdiadochokinesis",
  "HP:0001251 Ataxia" = "HP:0002078 Truncal ataxia",
  "HP:0001251 Ataxia" = "HP:0010867 Dyssynergia",
  "HP:0001251 Ataxia" = "HP:0002070 Limb Ataxia",
  )) |> count(hpo_terms, sort = T) |> filter (n >= 13)
ggplot(hpo_frequency, aes(x = n, y = fct_infreq(hpo_terms, (n)) |> fct_rev())) + geom_col(fill = "lightblue") + labs(title = "Most Frequent HPOs", subtitle = "Phenotypes with a frequency of 13+ across STR diseases", x = "Frequency", y = "HPO Term")

## FINAL HPO FREQUENCY PLOT
# Made using `fct_collapse()` function and manual matching with "children" from `ontologyIndex` data(hpo)
hpo_frequency <- unnested_hpo_terms |> mutate(
  hpo_terms = fct_collapse(hpo_terms,
  "HP:0001251 Ataxia" = c("HP:0001251 Ataxia", "HP:0001310 Dysmetria", "HP:0002066 Gait ataxia", "HP:0002070 Limb ataxia", "HP:0002073 Progressive cerebellar ataxia", "HP:0002075 Dysdiadochokinesis", "HP:0002078 Truncal ataxia", "HP:0010867 Dyssynergia", "HP:0002070 Limb Ataxia"),
  "HP:0001288 Gait disturbance" = c("HP:0001288 Gait disturbance", "HP:0002136 Broad-based gait", "HP:0002141 Gait imbalance", "HP:0002317 Unsteady gait", "HP:0002362 Shuffling gait", "HP:0002515 Waddling gait", "HP:0002527 Falls", "HP:0002540 Inability to walk", "HP:0003376 Steppage gait", "HP:0030051 Tip-toe gait", "HP:0031629 Impaired tandem gait"),
  "HP:0001260 Dysarthria" = c("HP:0001260 Dysarthria", "HP:0002464 Spastic dysarthria", "HP:0007024 Pseudobulbar paralysis", "HP:0008376 Nasal dysarthria"),
  "HP:0002015 Dysphagia" = c("HP:0002015 Dysphagia", "HP:0007024 Pseudobulbar paralysis", "HP:0031162 Impaired oropharyngeal swallow response", "HP:0200136 Oral-pharyngeal dysphagia"),
  "HP:0000639 Nystagmus" = c("HP:0000639 Nystagmus", "HP:0000640 Gaze-evoked nystagmus", "HP:0000666 Horizontal nystagmus", "HP:0010542 Vestibular nystagmus", "HP:0010544 Vertical nystagmus"),
  "HP:0001272 Cerebellar atrophy" = c("HP:0001272 Cerebellar atrophy", "HP:0006855 Cerebellar vermis atrophy", "HP:0007263 Spinocerebellar atrophy", "HP:0012082 Cerebellar Purkinje layer atrophy", "HP:0100275 Diffuse cerebellar atrophy"),
  "HP:0001249 Intellectual disability" = c("HP:0001249 Intellectual disability", "HP:0001256 Mild intellectual disability", "HP:0002342 Moderate intellectual disability", "HP:0006889 Borderline intellectual disability", "HP:0010864 Severe intellectual disability"),
  "HP:0001337 Tremor" = c("HP:0001337 Tremor", "HP:0002322 Resting tremor", "HP:0002345 Action tremor", "HP:0030188 Tremor by anatomical site"),
  "HP:0001250 Seizure" = c("HP:0001250 Seizure", "HP:0002069 Bilateral tonic-clonic seizure", "HP:0002133 Status epilepticus"),
  "HP:0001347 Hyperreflexia" = c("HP:0001347 Hyperreflexia", "HP:0001348 Brisk reflexes", "HP:0002169 Clonus", "HP:0006801 Hyperactive deep tendon reflexes"),
  "HP:0000726 Dementia" = c("HP:0000726 Dementia", "HP:0000727 Frontal lobe dementia", "HP:0002145 Frontotemporal dementia", "HP:0007123 Subcortical dementia"),
  "HP:0001265 Hyporeflexia" = ("HP:0001265 Hyporeflexia"),
  "HP:0002067 Bradykinesia" = ("HP:0002067 Bradykinesia"),
  "HP:0003487 Babinski sign" = ("HP:0003487 Babinski sign")
  )) |> count(hpo_terms, sort = T) |> filter (n >= 13)

ggplot(hpo_frequency, aes(x = n, y = fct_infreq(hpo_terms, (n)) |> fct_rev())) + geom_col(fill = "lightblue") + labs(title = "Most Frequent HPOs", subtitle = "Phenotypes with a frequency of 13+ across STR diseases", x = "Frequency", y = "HPO Term")

```

## Heatmapping with ontologyX Package Data

Using `onotologySimilarity` and `ontologyIndex` functions to create similarity matrices and distance clusters to then be visualized with `pheatmap`

```{r}
#| label: Similarity matrices, cluster dendrograms, and heat maps of HPO data


# Load packages and similarity matrix setup
library(pheatmap)
library(ontologyIndex)
library(ontologySimilarity)
data(hpo)
set.seed(1)
information_content <- descendants_IC(hpo)

# Create similarity matrix
  # First generate terms
  term_sets <- list(
  Ataxia = get_descendants(hpo, "HP:0001251"),
  Gait_Ataxia = get_descendants(hpo, "HP:0002066"),
  Gait_Disturbance = get_descendants(hpo, "HP:0001288"),
  Seizure = get_descendants(hpo, "HP:0001250")
  )
  # Next, calculate simialrity matrix
  sim_mat <- get_sim_grid(ontology = hpo, term_sets = term_sets)
  sim_mat
  # Convert to distance matrix (cluster dendrogram), which is necessary to visualize as   heat map 
  dist_mat <- max(sim_mat) - sim_mat
  plot(hclust(as.dist(dist_mat)))
  # Visualize as heatmap
   hclust_obj <- hclust(as.dist(dist_mat))
   pheatmap(sim_mat,
           cluster_rows = hclust_obj,
           cluster_cols = hclust_obj,
           show_rownames = TRUE,
           show_colnames = TRUE)
   
## BIG HPO TERMS HEATMAP
  # Manually input uncollapsed hpo_freq terms
  library(ontologyIndex)
  library(ontologySimilarity)
  data(hpo)
  set.seed(50)
  information_content <- descendants_IC(hpo)
# First, similarity matrix of individual terms  
  # Generate terms
  term_sets <- list(
    Dysarthria = "HP:0001260",
    Ataxia = "HP:0001251",
    Dysphagia = "HP:0002015",
    Cerebellar_Atrophy = "HP:0001272",
    Gait_Disturbance = "HP:0001288",
    Nystagmus = "HP:0000639",
    Intellectual_Disability = "HP:0001249",
    Tremor = "HP:0001337",
    Dementia = "HP:0000726",
    Seizure = "HP:0001250",
    Hyporeflexia = "HP:0001265",
    Hyperreflexia = "HP:0001347",
    Gait_Ataxia = "HP:0002066",
    Limb_Ataxia = "HP:0002070",
    Bradykinesia = "HP:0002067",
    Babinski_Sign = "HP:0003487",
    Progressive_Cerebellar_Ataxia = "HP:0002073",
    Depression = "HP:0000716",
    Hypotonia = "HP:0001252",
    Global_Developmental_Delay = "HP:0001263",
    Dysmetria = "HP:0001310",
    Muscle_Weakness = "HP:0001324",
    Dystonia = "HP:0001332",
    Myoclonus = "HP:0001336",
    Ptosis = "HP:0000508",
    Rigidity = "HP:0002063",
    Distal_Muscle_Weakness = "HP:0002460",
    Skeletal_Muscle_Atrophy = "HP:0003202",
    Spasticity = "HP:0001257",
    Parkinsonism = "HP:0001300"
  )
  # Similarity matrix
  sim_mat <- get_sim_grid(ontology = hpo, term_sets = term_sets)
  sim_mat
  # Cluster dendrogram
  dist_mat <- max(sim_mat) - sim_mat
  plot(hclust(as.dist(dist_mat)))
  # Heat map
  hclust_obj <- hclust(as.dist(dist_mat))
  pheatmap(sim_mat,
           cluster_rows = hclust_obj,
           cluster_cols = hclust_obj,
           show_rownames = TRUE,
           show_colnames = TRUE,
           fontsize = 6.5
           )


## DISEASE CLUSTERS HEATMAP!
library(jsonlite)
library(tidyverse)
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)
disease_hpo <- loci |> select(c(disease_id, hpo_terms)) |> unnest_longer(hpo_terms) |> mutate(hpo_id = str_extract(hpo_terms, "^HP:\\d+")) |> select(disease_id, hpo_id) |> group_by(disease_id) |> nest(hpo_id = hpo_id) |> rowwise() |> mutate(hpo_id = paste(hpo_id, collapse = ', ')) |> ungroup()
library(ontologyIndex)
library(ontologySimilarity)
data(hpo)
set.seed(50)
information_content <- descendants_IC(hpo)
library(pheatmap)
  # Generate terms (used 'str_view(disease_hpo$hpo_id[])' to copy    HPO terms) --> Similarity matrix --> Heat map
  term_set <- list(
  OPDM5 = c("HP:0000183", "HP:0000218", "HP:0000301", "HP:0000408", "HP:0000590", "HP:0000597", "HP:0001284", "HP:0001288", "HP:0001604", "HP:0001824", "HP:0002058", "HP:0002091", "HP:0002100", "HP:0002505", "HP:0002705", "HP:0002747", "HP:0007149", "HP:0007838", "HP:0008376", "HP:0008756", "HP:0008944", "HP:0008959", "HP:0008963", "HP:0008997", "HP:0009027", "HP:0009053", "HP:0009063", "HP:0009073", "HP:0010550", "HP:0030192", "HP:0030319", "HP:0031162", "HP:0200136", "HP:0430015", "HP:3000005", "HP:3000010"),
  FRAXE = c("HP:0000252", "HP:0000256", "HP:0000286", "HP:0000426", "HP:0000713", "HP:0000718", "HP:0000722", "HP:0000729", "HP:0000750", "HP:0000752", "HP:0001249", "HP:0001328", "HP:0001511", "HP:0001609", "HP:0002311", "HP:0002312", "HP:0004209", "HP:0004322", "HP:0009904", "HP:0011341", "HP:0012172", "HP:0012471", "HP:0025116", "HP:0100023", "HP:0100710"),
  SBMA = c("HP:0000029", "HP:0000144", "HP:0000153", "HP:0000763", "HP:0000771", "HP:0001252", "HP:0001260", "HP:0001265", "HP:0001283", "HP:0001288", "HP:0001337", "HP:0001618", "HP:0002015", "HP:0002380", "HP:0003119", "HP:0003202", "HP:0003236", "HP:0003394", "HP:0003560", "HP:0003690", "HP:0005978", "HP:0008981", "HP:0009830", "HP:0100022", "HP:0100639"),
  EIEE1 = c("HP:0000054", "HP:0000252", "HP:0000568", "HP:0000817", "HP:0001249", "HP:0001263", "HP:0001266", "HP:0001276", "HP:0001285", "HP:0001332", "HP:0001347", "HP:0001357", "HP:0001510", "HP:0002015", "HP:0002094", "HP:0002119", "HP:0002123", "HP:0002188", "HP:0002283", "HP:0002421", "HP:0002521", "HP:0007256", "HP:0007359", "HP:0008936", "HP:0010851", "HP:0011153", "HP:0011344", "HP:0012469", "HP:0025357", "HP:0032792", "HP:0100660", "HP:0200134"),
  PRTS = c("HP:0000053", "HP:0000325", "HP:0000708", "HP:0000750", "HP:0001249", "HP:0001250", "HP:0001256", "HP:0001260", "HP:0001288", "HP:0001371", "HP:0002061", "HP:0002342", "HP:0002353", "HP:0002451", "HP:0004373", "HP:0007380", "HP:0012385", "HP:0012469"),
  DRPLA = c("HP:0000597", "HP:0000639", "HP:0000643", "HP:0000726", "HP:0001138", "HP:0001152", "HP:0001249", "HP:0001250", "HP:0001251", "HP:0001260", "HP:0001265", "HP:0001266", "HP:0001300", "HP:0001310", "HP:0001332", "HP:0001336", "HP:0002066", "HP:0002070", "HP:0002072", "HP:0002073", "HP:0002075", "HP:0002078", "HP:0002172", "HP:0002345", "HP:0002354", "HP:0004305", "HP:0007047", "HP:0010831", "HP:0010867", "HP:0012048", "HP:0030890", "HP:0100543"),
  SCA1 = c("HP:0000496", "HP:0000514", "HP:0000543", "HP:0000597", "HP:0000623", "HP:0000639", "HP:0000640", "HP:0000641", "HP:0000648", "HP:0001151", "HP:0001252", "HP:0001257", "HP:0001260", "HP:0001265", "HP:0001272", "HP:0001283", "HP:0001284", "HP:0001288", "HP:0001290", "HP:0001310", "HP:0001324", "HP:0001332", "HP:0001347", "HP:0001350", "HP:0002015", "HP:0002067", "HP:0002070", "HP:0002071", "HP:0002072", "HP:0002073", "HP:0002075", "HP:0002078", "HP:0002141", "HP:0002168", "HP:0002174", "HP:0002198", "HP:0002354", "HP:0002363", "HP:0002380", "HP:0002460", "HP:0002483", "HP:0002495", "HP:0002503", "HP:0002542", "HP:0002839", "HP:0002878", "HP:0003202", "HP:0003394", "HP:0003401", "HP:0003431", "HP:0003448", "HP:0003487", "HP:0003693", "HP:0003701", "HP:0006801", "HP:0006937", "HP:0007001", "HP:0007006", "HP:0007078", "HP:0007263", "HP:0007328", "HP:0007338", "HP:0007366", "HP:0007377", "HP:0007928", "HP:0009830", "HP:0010831", "HP:0025331", "HP:0025401", "HP:0030216", "HP:0040129", "HP:0100543", "HP:0410011"),
  SCA10 = c("HP:0000012", "HP:0000020", "HP:0000639", "HP:0000640", "HP:0000716", "HP:0000718", "HP:0000726", "HP:0000741", "HP:0000762", "HP:0001250", "HP:0001260", "HP:0001265", "HP:0001271", "HP:0001272", "HP:0001290", "HP:0001310", "HP:0001347", "HP:0002015", "HP:0002061", "HP:0002062", "HP:0002066", "HP:0002067", "HP:0002070", "HP:0002071", "HP:0002073", "HP:0002075", "HP:0002080", "HP:0002133", "HP:0002141", "HP:0002168", "HP:0002197", "HP:0002311", "HP:0002317", "HP:0002360", "HP:0002384", "HP:0002936", "HP:0003487", "HP:0007256", "HP:0007289", "HP:0007772", "HP:0011153", "HP:0011198", "HP:0030186", "HP:0100660"),
  SCA2 = c("HP:0000020", "HP:0000510", "HP:0000514", "HP:0000597", "HP:0000602", "HP:0000623", "HP:0000639", "HP:0000640", "HP:0000641", "HP:0000657", "HP:0000726", "HP:0001151", "HP:0001251", "HP:0001252", "HP:0001257", "HP:0001260", "HP:0001265", "HP:0001272", "HP:0001290", "HP:0001300", "HP:0001310", "HP:0001332", "HP:0001336", "HP:0002015", "HP:0002063", "HP:0002066", "HP:0002067", "HP:0002070", "HP:0002072", "HP:0002073", "HP:0002075", "HP:0002120", "HP:0002172", "HP:0002174", "HP:0002198", "HP:0002317", "HP:0002345", "HP:0002380", "HP:0002495", "HP:0002503", "HP:0002536", "HP:0002542", "HP:0002839", "HP:0003133", "HP:0003394", "HP:0003487", "HP:0003693", "HP:0006801", "HP:0006955", "HP:0008311", "HP:0012082", "HP:0012762", "HP:0025461", "HP:0030186", "HP:0045007"),
  SCA3_MJD = c("HP:0000508", "HP:0000520", "HP:0000544", "HP:0000590", "HP:0000623", "HP:0000639", "HP:0000640", "HP:0000641", "HP:0000651", "HP:0000726", "HP:0000750", "HP:0001151", "HP:0001251", "HP:0001257", "HP:0001260", "HP:0001272", "HP:0001300", "HP:0001332", "HP:0001347", "HP:0001605", "HP:0001751", "HP:0002015", "HP:0002063", "HP:0002067", "HP:0002070", "HP:0002071", "HP:0002073", "HP:0002078", "HP:0002171", "HP:0002172", "HP:0002198", "HP:0002312", "HP:0002380", "HP:0002495", "HP:0002503", "HP:0002839", "HP:0003202", "HP:0003394", "HP:0003438", "HP:0003487", "HP:0003693", "HP:0004370", "HP:0007089", "HP:0007256", "HP:0012332", "HP:0012532", "HP:0030454"),
  SCA7 = c("HP:0000514", "HP:0000529", "HP:0000548", "HP:0000572", "HP:0000580", "HP:0000597", "HP:0000602", "HP:0000608", "HP:0000613", "HP:0000618", "HP:0000623", "HP:0000639", "HP:0000648", "HP:0000709", "HP:0001098", "HP:0001251", "HP:0001257", "HP:0001260", "HP:0001263", "HP:0001268", "HP:0001270", "HP:0001272", "HP:0001310", "HP:0001319", "HP:0001324", "HP:0001337", "HP:0001347", "HP:0001508", "HP:0001635", "HP:0002015", "HP:0002059", "HP:0002071", "HP:0002072", "HP:0002073", "HP:0002075", "HP:0002310", "HP:0002542", "HP:0003474", "HP:0003487", "HP:0007663", "HP:0011968", "HP:0012047", "HP:0012452"),
  SCA8 = c("HP:0000020", "HP:0000273", "HP:0000514", "HP:0000639", "HP:0000641", "HP:0000716", "HP:0000763", "HP:0000802", "HP:0001251", "HP:0001257", "HP:0001260", "HP:0001272", "HP:0001332", "HP:0001337", "HP:0001347", "HP:0002015", "HP:0002062", "HP:0002063", "HP:0002066", "HP:0002067", "HP:0002070", "HP:0002073", "HP:0002172", "HP:0002311", "HP:0002317", "HP:0002464", "HP:0002495", "HP:0002835", "HP:0006855", "HP:0007256", "HP:0007772", "HP:0009830", "HP:0012110"),
  SCA31 = c("HP:0000365", "HP:0000407", "HP:0000639", "HP:0001251", "HP:0001257", "HP:0001260", "HP:0001265", "HP:0001272", "HP:0001337", "HP:0001347", "HP:0002066", "HP:0002070", "HP:0002495", "HP:0006801", "HP:0007979"),
  FTDALS1 = c("HP:0000605", "HP:0000716", "HP:0000726", "HP:0000738", "HP:0000741", "HP:0000746", "HP:0001260", "HP:0001300", "HP:0001324", "HP:0002059", "HP:0002145", "HP:0002171", "HP:0002186", "HP:0002273", "HP:0002366", "HP:0002385", "HP:0002442", "HP:0002529", "HP:0003202", "HP:0007308", "HP:0007354"),
  SCA6 = c("HP:0000504", "HP:0000639", "HP:0000643", "HP:0000651", "HP:0000763", "HP:0001251", "HP:0001260", "HP:0001272", "HP:0001310", "HP:0001332", "HP:0001347", "HP:0001350", "HP:0002015", "HP:0002017", "HP:0002066", "HP:0002073", "HP:0002078", "HP:0002080", "HP:0002083", "HP:0002120", "HP:0002172", "HP:0002311", "HP:0002317", "HP:0002321", "HP:0002359", "HP:0002505", "HP:0003487", "HP:0007670", "HP:0007772", "HP:0007979", "HP:0010544", "HP:0030511", "HP:0030842"),
  JBS = c("HP:0000003", "HP:0000023", "HP:0000028", "HP:0000047", "HP:0000060", "HP:0000066", "HP:0000126", "HP:0000174", "HP:0000238", "HP:0000243", "HP:0000252", "HP:0000256", "HP:0000286", "HP:0000316", "HP:0000319", "HP:0000324", "HP:0000343", "HP:0000347", "HP:0000348", "HP:0000358", "HP:0000369", "HP:0000431", "HP:0000463", "HP:0000465", "HP:0000470", "HP:0000482", "HP:0000486", "HP:0000494", "HP:0000499", "HP:0000506", "HP:0000508", "HP:0000518", "HP:0000567", "HP:0000568", "HP:0000579", "HP:0000612", "HP:0000625", "HP:0000646", "HP:0000648", "HP:0000656", "HP:0000767", "HP:0000921", "HP:0000964", "HP:0001104", "HP:0001156", "HP:0001161", "HP:0001249", "HP:0001250", "HP:0001257", "HP:0001263", "HP:0001274", "HP:0001290", "HP:0001302", "HP:0001360", "HP:0001371", "HP:0001508", "HP:0001510", "HP:0001511", "HP:0001622", "HP:0001629", "HP:0001631", "HP:0001650", "HP:0001680", "HP:0001734", "HP:0001763", "HP:0001770", "HP:0001831", "HP:0001847", "HP:0001863", "HP:0001873", "HP:0001883", "HP:0002007", "HP:0002019", "HP:0002021", "HP:0002059", "HP:0002119", "HP:0002205", "HP:0002247", "HP:0002414", "HP:0002566", "HP:0002650", "HP:0002827", "HP:0003196", "HP:0003312", "HP:0004209", "HP:0004322", "HP:0004378", "HP:0004383", "HP:0004397", "HP:0005280", "HP:0005469", "HP:0005528", "HP:0006101", "HP:0007018", "HP:0007302", "HP:0008872", "HP:0008947", "HP:0009906", "HP:0010059", "HP:0010761", "HP:0010806", "HP:0100753", "HP:0100840"),
  MODY8 = c("HP:0001738", "HP:0002027", "HP:0003074", "HP:0004904", "HP:0040217"),
  DM2 = c("HP:0000026", "HP:0000135", "HP:0000407", "HP:0000518", "HP:0000798", "HP:0000975", "HP:0001249", "HP:0001262", "HP:0001265", "HP:0001348", "HP:0001638", "HP:0001649", "HP:0001962", "HP:0002015", "HP:0002019", "HP:0002027", "HP:0002292", "HP:0002360", "HP:0002486", "HP:0002850", "HP:0002870", "HP:0002926", "HP:0003077", "HP:0003202", "HP:0003236", "HP:0003326", "HP:0003327", "HP:0003552", "HP:0003554", "HP:0003700", "HP:0003701", "HP:0003722", "HP:0004313", "HP:0004315", "HP:0005978", "HP:0006682", "HP:0007787", "HP:0007889", "HP:0008189", "HP:0008232", "HP:0008981", "HP:0011712", "HP:0012036", "HP:0012378", "HP:0012452", "HP:0012899", "HP:0030319", "HP:0030891", "HP:0031546", "HP:0100543"),
  EDM1_PSACH = c("HP:0000763", "HP:0000926", "HP:0001156", "HP:0001249", "HP:0001288", "HP:0001376", "HP:0001377", "HP:0001382", "HP:0001385", "HP:0001387", "HP:0001498", "HP:0001763", "HP:0002341", "HP:0002515", "HP:0002650", "HP:0002656", "HP:0002663", "HP:0002758", "HP:0002761", "HP:0002808", "HP:0002812", "HP:0002816", "HP:0002829", "HP:0002834", "HP:0002857", "HP:0002938", "HP:0002970", "HP:0003015", "HP:0003016", "HP:0003025", "HP:0003026", "HP:0003049", "HP:0003090", "HP:0003093", "HP:0003170", "HP:0003180", "HP:0003300", "HP:0003301", "HP:0003311", "HP:0003312", "HP:0003365", "HP:0003414", "HP:0003498", "HP:0003502", "HP:0003510", "HP:0003756", "HP:0004019", "HP:0004042", "HP:0004236", "HP:0004568", "HP:0005063", "HP:0005720", "HP:0005743", "HP:0006094", "HP:0006149", "HP:0006429", "HP:0006460", "HP:0006467", "HP:0006499", "HP:0008800", "HP:0008807", "HP:0008833", "HP:0008839", "HP:0008843", "HP:0008873", "HP:0009107", "HP:0009487", "HP:0009803", "HP:0009826", "HP:0009882", "HP:0010049", "HP:0010236", "HP:0010579", "HP:0010582", "HP:0010585", "HP:0010646", "HP:0011405", "HP:0012307", "HP:0020152", "HP:0030839", "HP:0030840", "HP:0030973", "HP:0045086", "HP:0100168", "HP:0100531", "HP:0100864"),
  EPM_DEE = c("HP:0000726", "HP:0000992", "HP:0001249", "HP:0001251", "HP:0001260", "HP:0001336", "HP:0002070", "HP:0002080", "HP:0002392", "HP:0007000"),
  EPM1 = c("HP:0000726", "HP:0000992", "HP:0001249", "HP:0001251", "HP:0001256", "HP:0001260", "HP:0001268", "HP:0001336", "HP:0002069", "HP:0002070", "HP:0002080", "HP:0002121", "HP:0002392", "HP:0007000", "HP:0010850", "HP:0011182"),
  SCA37 = c("HP:0000407", "HP:0000549", "HP:0000639", "HP:0000666", "HP:0001251", "HP:0001260", "HP:0001272", "HP:0001288", "HP:0001336", "HP:0001337", "HP:0002015", "HP:0002075", "HP:0002078", "HP:0002167", "HP:0002168", "HP:0002317", "HP:0002359", "HP:0002396", "HP:0002406", "HP:0002527", "HP:0003474", "HP:0006855", "HP:0100275"),
  FRA12A = c("HP:0000708", "HP:0000962", "HP:0001019", "HP:0001249", "HP:0001250", "HP:0001263", "HP:0002783"),
  DMD = c("HP:0000750", "HP:0001252", "HP:0001256", "HP:0001263", "HP:0001265", "HP:0001270", "HP:0001324", "HP:0001328", "HP:0001371", "HP:0001635", "HP:0001638", "HP:0001644", "HP:0001771", "HP:0002091", "HP:0002093", "HP:0002194", "HP:0002505", "HP:0002515", "HP:0002650", "HP:0002747", "HP:0002791", "HP:0002870", "HP:0002878", "HP:0003089", "HP:0003115", "HP:0003202", "HP:0003236", "HP:0003307", "HP:0003323", "HP:0003391", "HP:0003551", "HP:0003560", "HP:0003701", "HP:0003707", "HP:0006380", "HP:0008981", "HP:0011675", "HP:0030051", "HP:0100543"),
  DM1 = c("HP:0000026", "HP:0000029", "HP:0000135", "HP:0000144", "HP:0000467", "HP:0000483", "HP:0000518", "HP:0000540", "HP:0000602", "HP:0000708", "HP:0000712", "HP:0000716", "HP:0000717", "HP:0000718", "HP:0000729", "HP:0000736", "HP:0000739", "HP:0000802", "HP:0000815", "HP:0000819", "HP:0000824", "HP:0000842", "HP:0000855", "HP:0000867", "HP:0001081", "HP:0001249", "HP:0001252", "HP:0001256", "HP:0001260", "HP:0001262", "HP:0001263", "HP:0001268", "HP:0001288", "HP:0001290", "HP:0001319", "HP:0001324", "HP:0001328", "HP:0001349", "HP:0001488", "HP:0001558", "HP:0001561", "HP:0001596", "HP:0001644", "HP:0001762", "HP:0002014", "HP:0002015", "HP:0002019", "HP:0002059", "HP:0002093", "HP:0002098", "HP:0002120", "HP:0002234", "HP:0002292", "HP:0002460", "HP:0002486", "HP:0002494", "HP:0002500", "HP:0002527", "HP:0002540", "HP:0002747", "HP:0002870", "HP:0002878", "HP:0002910", "HP:0002926", "HP:0003003", "HP:0003124", "HP:0003202", "HP:0003326", "HP:0003477", "HP:0003547", "HP:0003693", "HP:0003701", "HP:0003722", "HP:0003740", "HP:0003749", "HP:0004389", "HP:0004749", "HP:0004755", "HP:0004887", "HP:0005110", "HP:0006677", "HP:0006889", "HP:0007010", "HP:0007663", "HP:0007787", "HP:0007941", "HP:0008069", "HP:0008770", "HP:0008872", "HP:0009027", "HP:0009113", "HP:0009830", "HP:0010794", "HP:0010804", "HP:0010864", "HP:0010952", "HP:0011470", "HP:0011705", "HP:0011999", "HP:0012054", "HP:0012114", "HP:0012248", "HP:0012378", "HP:0012899", "HP:0012901", "HP:0012903", "HP:0025169", "HP:0025318", "HP:0030192", "HP:0030319", "HP:0030692", "HP:0031466", "HP:0031546", "HP:0040171", "HP:0040173", "HP:0040198", "HP:0100284", "HP:0100543", "HP:0200136", "HP:0410011"),
  RCPS = c("HP:0000160", "HP:0000175", "HP:0000193", "HP:0000201", "HP:0000218", "HP:0000308", "HP:0000347", "HP:0000369", "HP:0000377", "HP:0000411", "HP:0000448", "HP:0000690", "HP:0000750", "HP:0001245", "HP:0001263", "HP:0001328", "HP:0001608", "HP:0001609", "HP:0001762", "HP:0002643", "HP:0002827", "HP:0002984", "HP:0003038", "HP:0004209", "HP:0004322", "HP:0004987", "HP:0005011", "HP:0005736", "HP:0005867", "HP:0006355", "HP:0006536", "HP:0008744", "HP:0008753", "HP:0008807", "HP:0008872", "HP:0009094", "HP:0009237", "HP:0009486", "HP:0009601", "HP:0009623", "HP:0009778", "HP:0009803", "HP:0010049", "HP:0010109", "HP:0010487", "HP:0010565", "HP:0010752", "HP:0011051", "HP:0011968", "HP:0012789", "HP:0030043", "HP:0100499", "HP:0200154"),
  SCA27B = c("HP:0000651", "HP:0001260", "HP:0001272", "HP:0002066", "HP:0002070", "HP:0002174", "HP:0002321", "HP:0007979", "HP:0010545"),
  FXS_FXTAS_POF1 = c("HP:0000020", "HP:0000053", "HP:0000246", "HP:0000256", "HP:0000275", "HP:0000276", "HP:0000280", "HP:0000298", "HP:0000303", "HP:0000365", "HP:0000388", "HP:0000389", "HP:0000400", "HP:0000411", "HP:0000486", "HP:0000639", "HP:0000713", "HP:0000716", "HP:0000717", "HP:0000722", "HP:0000726", "HP:0000734", "HP:0000737", "HP:0000739", "HP:0000750", "HP:0000752", "HP:0000767", "HP:0000802", "HP:0000817", "HP:0000821", "HP:0000822", "HP:0000837", "HP:0000858", "HP:0001152", "HP:0001250", "HP:0001251", "HP:0001252", "HP:0001260", "HP:0001265", "HP:0001268", "HP:0001272", "HP:0001288", "HP:0001300", "HP:0001310", "HP:0001324", "HP:0001382", "HP:0001634", "HP:0001763", "HP:0002003", "HP:0002007", "HP:0002015", "HP:0002020", "HP:0002050", "HP:0002063", "HP:0002066", "HP:0002067", "HP:0002075", "HP:0002080", "HP:0002120", "HP:0002167", "HP:0002172", "HP:0002174", "HP:0002194", "HP:0002322", "HP:0002342", "HP:0002345", "HP:0002354", "HP:0002360", "HP:0002363", "HP:0002457", "HP:0002506", "HP:0002607", "HP:0002615", "HP:0002650", "HP:0002839", "HP:0003326", "HP:0003564", "HP:0004970", "HP:0006099", "HP:0006886", "HP:0007010", "HP:0007018", "HP:0007165", "HP:0007340", "HP:0008209", "HP:0008640", "HP:0008770", "HP:0009830", "HP:0012083", "HP:0012169", "HP:0012332", "HP:0012534", "HP:0030216", "HP:0031629", "HP:0100023", "HP:0100275", "HP:0100515", "HP:0100716"),
  BPES = c("HP:0000013", "HP:0000141", "HP:0000218", "HP:0000252", "HP:0000286", "HP:0000378", "HP:0000431", "HP:0000482", "HP:0000486", "HP:0000506", "HP:0000508", "HP:0000537", "HP:0000540", "HP:0000545", "HP:0000568", "HP:0000581", "HP:0000639", "HP:0000664", "HP:0000769", "HP:0000837", "HP:0000858", "HP:0002225", "HP:0002553", "HP:0005280", "HP:0008209", "HP:0008222", "HP:0011481", "HP:0045025"),
  FRDA = c("HP:0000505", "HP:0000639", "HP:0000648", "HP:0000649", "HP:0000763", "HP:0000819", "HP:0001123", "HP:0001251", "HP:0001260", "HP:0001626", "HP:0001635", "HP:0001639", "HP:0001761", "HP:0002066", "HP:0002070", "HP:0002495", "HP:0002522", "HP:0002650", "HP:0003115", "HP:0003209", "HP:0003232", "HP:0003448", "HP:0003487", "HP:0007078", "HP:0007663", "HP:0010831"),
  OPDM2 = c("HP:0000508", "HP:0000544", "HP:0001283", "HP:0002460", "HP:0003236", "HP:0003458", "HP:0003557", "HP:0003805", "HP:0012548", "HP:0030319", "HP:0100297"),
  aFTLDU = "HP:0002145",
  HFGI = c("HP:0000010", "HP:0000011", "HP:0000013", "HP:0000041", "HP:0000047", "HP:0000048", "HP:0000054", "HP:0000074", "HP:0000076", "HP:0000083", "HP:0000126", "HP:0000130", "HP:0000486", "HP:0000787", "HP:0000795", "HP:0000807", "HP:0000808", "HP:0000813", "HP:0000960", "HP:0001032", "HP:0001081", "HP:0001162", "HP:0001216", "HP:0001241", "HP:0001245", "HP:0001629", "HP:0001762", "HP:0001763", "HP:0001773", "HP:0001792", "HP:0001798", "HP:0001799", "HP:0001822", "HP:0001857", "HP:0001885", "HP:0002019", "HP:0002144", "HP:0003259", "HP:0003762", "HP:0004180", "HP:0004209", "HP:0004220", "HP:0004243", "HP:0005048", "HP:0006110", "HP:0007477", "HP:0008080", "HP:0008103", "HP:0008111", "HP:0008368", "HP:0008551", "HP:0008740", "HP:0009237", "HP:0009246", "HP:0009290", "HP:0009464", "HP:0009566", "HP:0009623", "HP:0009702", "HP:0009778", "HP:0009882", "HP:0010034", "HP:0010055", "HP:0010086", "HP:0010103", "HP:0010105", "HP:0010109", "HP:0010584", "HP:0011937", "HP:0012330", "HP:0012386", "HP:0012704", "HP:0025633", "HP:0030084", "HP:0040025", "HP:0100329", "HP:0100341", "HP:0100343", "HP:0100347", "HP:0100499", "HP:0200055", "HP:6000184"),
  HFGII = c("HP:0000010", "HP:0000011", "HP:0000013", "HP:0000041", "HP:0000047", "HP:0000048", "HP:0000054", "HP:0000074", "HP:0000076", "HP:0000083", "HP:0000126", "HP:0000130", "HP:0000486", "HP:0000787", "HP:0000795", "HP:0000807", "HP:0000808", "HP:0000813", "HP:0000960", "HP:0001032", "HP:0001081", "HP:0001162", "HP:0001216", "HP:0001241", "HP:0001245", "HP:0001629", "HP:0001762", "HP:0001763", "HP:0001773", "HP:0001792", "HP:0001798", "HP:0001799", "HP:0001822", "HP:0001857", "HP:0001885", "HP:0002019", "HP:0002144", "HP:0003259", "HP:0003762", "HP:0004180", "HP:0004209", "HP:0004220", "HP:0004243", "HP:0005048", "HP:0006110", "HP:0007477", "HP:0008080", "HP:0008103", "HP:0008111", "HP:0008368", "HP:0008551", "HP:0008740", "HP:0009237", "HP:0009246", "HP:0009290", "HP:0009464", "HP:0009566", "HP:0009623", "HP:0009702", "HP:0009778", "HP:0009882", "HP:0010034", "HP:0010055", "HP:0010086", "HP:0010103", "HP:0010105", "HP:0010109", "HP:0010584", "HP:0011937", "HP:0012330", "HP:0012386", "HP:0012704", "HP:0025633", "HP:0030084", "HP:0040025", "HP:0100329", "HP:0100341", "HP:0100343", "HP:0100347", "HP:0100499", "HP:0200055", "HP:6000184"),
  HFGIII = c("HP:0000010", "HP:0000011", "HP:0000013", "HP:0000041", "HP:0000047", "HP:0000048", "HP:0000054", "HP:0000074", "HP:0000076", "HP:0000083", "HP:0000126", "HP:0000130", "HP:0000486", "HP:0000787", "HP:0000795", "HP:0000807", "HP:0000808", "HP:0000813", "HP:0000960", "HP:0001032", "HP:0001081", "HP:0001162", "HP:0001216", "HP:0001241", "HP:0001245", "HP:0001629", "HP:0001762", "HP:0001763", "HP:0001773", "HP:0001792", "HP:0001798", "HP:0001799", "HP:0001822", "HP:0001857", "HP:0001885", "HP:0002019", "HP:0002144", "HP:0003259", "HP:0003762", "HP:0004180", "HP:0004209", "HP:0004220", "HP:0004243", "HP:0005048", "HP:0006110", "HP:0007477", "HP:0008080", "HP:0008103", "HP:0008111", "HP:0008368", "HP:0008551", "HP:0008740", "HP:0009237", "HP:0009246", "HP:0009290", "HP:0009464", "HP:0009566", "HP:0009623", "HP:0009702", "HP:0009778", "HP:0009882", "HP:0010034", "HP:0010055", "HP:0010086", "HP:0010103", "HP:0010105", "HP:0010109", "HP:0010584", "HP:0011937", "HP:0012330", "HP:0012386", "HP:0012704", "HP:0025633", "HP:0030084", "HP:0040025", "HP:0100329", "HP:0100341", "HP:0100343", "HP:0100347", "HP:0100499", "HP:0200055", "HP:6000184"),
  SD5 = c("HP:0001501", "HP:0001830", "HP:0001841", "HP:0004209", "HP:0004220", "HP:0004692", "HP:0006042", "HP:0006101", "HP:0006159", "HP:0008083", "HP:0009185", "HP:0010055", "HP:0011939"),
  HD = c("HP:0000496", "HP:0000657", "HP:0000713", "HP:0000716", "HP:0000718", "HP:0000722", "HP:0000726", "HP:0000734", "HP:0000737", "HP:0000738", "HP:0000739", "HP:0000741", "HP:0000746", "HP:0000751", "HP:0001250", "HP:0001262", "HP:0001268", "HP:0001272", "HP:0001288", "HP:0001332", "HP:0001336", "HP:0001347", "HP:0001824", "HP:0002059", "HP:0002063", "HP:0002066", "HP:0002067", "HP:0002072", "HP:0002141", "HP:0002169", "HP:0002171", "HP:0002300", "HP:0002312", "HP:0002340", "HP:0002354", "HP:0002375", "HP:0002500", "HP:0002529", "HP:0002540", "HP:0002591", "HP:0003107", "HP:0003324", "HP:0003487", "HP:0004305", "HP:0004408", "HP:0007010", "HP:0009088", "HP:0010794", "HP:0025401", "HP:0030842", "HP:0030955", "HP:0031473", "HP:0031589", "HP:0031843", "HP:0031845", "HP:0040140", "HP:0045082", "HP:0100785", "HP:0200136"),
  HDL2 = c("HP:0000708", "HP:0000716", "HP:0000726", "HP:0000737", "HP:0000738", "HP:0000739", "HP:0000741", "HP:0000746", "HP:0000751", "HP:0001260", "HP:0001288", "HP:0001300", "HP:0001332", "HP:0001347", "HP:0001824", "HP:0002060", "HP:0002063", "HP:0002067", "HP:0002072", "HP:0002120", "HP:0002340", "HP:0002345", "HP:0002354", "HP:0002476", "HP:0004302", "HP:0004305", "HP:0007123", "HP:0010994", "HP:0030216", "HP:0100022"),
  OPDM1 = c("HP:0000218", "HP:0000407", "HP:0000508", "HP:0000544", "HP:0000597", "HP:0001251", "HP:0001260", "HP:0001284", "HP:0001288", "HP:0001324", "HP:0001337", "HP:0001488", "HP:0001611", "HP:0001618", "HP:0001639", "HP:0001644", "HP:0001824", "HP:0002015", "HP:0002058", "HP:0002091", "HP:0002098", "HP:0002460", "HP:0002500", "HP:0002747", "HP:0002835", "HP:0003236", "HP:0003394", "HP:0003458", "HP:0003557", "HP:0003693", "HP:0003701", "HP:0003736", "HP:0003805", "HP:0004757", "HP:0008756", "HP:0009027", "HP:0010628", "HP:0012416", "HP:0012444", "HP:0032341", "HP:0100284"),
  FAME3 = c("HP:0001288", "HP:0001312", "HP:0001336", "HP:0001337", "HP:0001340", "HP:0001351", "HP:0002069", "HP:0007359", "HP:0010852"),
  CHNG3 = c("HP:0000821", "HP:0002925", "HP:0005987", "HP:0025484"),
  ADTKD = c("HP:0000089", "HP:0000092", "HP:0000096", "HP:0000108", "HP:0000127", "HP:0000822", "HP:0001903", "HP:0001970", "HP:0001997", "HP:0002048", "HP:0002120", "HP:0002149", "HP:0002615", "HP:0003259", "HP:0003774", "HP:0004732", "HP:0005576", "HP:0005583", "HP:0012213"),
  NME = c("HP:0000473", "HP:0000486", "HP:0000511", "HP:0000639", "HP:0000711", "HP:0000737", "HP:0001250", "HP:0001251", "HP:0001252", "HP:0001259", "HP:0001260", "HP:0001263", "HP:0001298", "HP:0001337", "HP:0002063", "HP:0002119", "HP:0002151", "HP:0002181", "HP:0002196", "HP:0002273", "HP:0002283", "HP:0002318", "HP:0002352", "HP:0002376", "HP:0002490", "HP:0002878", "HP:0003128", "HP:0006897", "HP:0011922", "HP:0012378", "HP:0012444", "HP:0012706", "HP:0012762", "HP:0030915", "HP:0032794", "HP:0046507", "HP:0200041"),
  ALS1 = c("HP:0001257", "HP:0001260", "HP:0001324", "HP:0001347", "HP:0002015", "HP:0002314", "HP:0002380", "HP:0002398", "HP:0003202", "HP:0003394", "HP:0007024", "HP:0007354", "HP:0010535"),
  SCA36 = c("HP:0000365", "HP:0000508", "HP:0000511", "HP:0000514", "HP:0000622", "HP:0000639", "HP:0000651", "HP:0001251", "HP:0001252", "HP:0001260", "HP:0001272", "HP:0001276", "HP:0001288", "HP:0001308", "HP:0001310", "HP:0001324", "HP:0001347", "HP:0002015", "HP:0002066", "HP:0002070", "HP:0002076", "HP:0002078", "HP:0002080", "HP:0002311", "HP:0002321", "HP:0002346", "HP:0002378", "HP:0002380", "HP:0002607", "HP:0003202", "HP:0003445", "HP:0003487", "HP:0007001", "HP:0007018", "HP:0007772", "HP:0012473", "HP:0045084"),
  NIID = c("HP:0000020", "HP:0000600", "HP:0000602", "HP:0000616", "HP:0000639", "HP:0000648", "HP:0000708", "HP:0000726", "HP:0001250", "HP:0001251", "HP:0001260", "HP:0001265", "HP:0001276", "HP:0001279", "HP:0001288", "HP:0001324", "HP:0001337", "HP:0001347", "HP:0002063", "HP:0002119", "HP:0002167", "HP:0002352", "HP:0002353", "HP:0002572", "HP:0002650", "HP:0002922", "HP:0003298", "HP:0003312", "HP:0003403", "HP:0003431", "HP:0003448", "HP:0003457", "HP:0003474", "HP:0007185", "HP:0012229", "HP:0100022", "HP:0100543"),
  OPML1 = c("HP:0000508", "HP:0000544", "HP:0001251", "HP:0001260", "HP:0001337", "HP:0002015", "HP:0002059", "HP:0002460", "HP:0002579", "HP:0002878", "HP:0003701", "HP:0030319"),
  OPMD = c("HP:0000298", "HP:0000467", "HP:0000508", "HP:0001260", "HP:0001288", "HP:0001611", "HP:0002015", "HP:0002460", "HP:0003200", "HP:0003236", "HP:0003690", "HP:0003701", "HP:0007838", "HP:0010628"),
  CCHS = c("HP:0000358", "HP:0000369", "HP:0000494", "HP:0000975", "HP:0001250", "HP:0001252", "HP:0002093", "HP:0002104", "HP:0002251", "HP:0002270", "HP:0002791", "HP:0002877", "HP:0003005", "HP:0003006", "HP:0004370", "HP:0006747", "HP:0007110", "HP:0011968", "HP:0012332", "HP:0012416", "HP:0012418", "HP:0012450", "HP:0031857", "HP:0031861", "HP:0100006", "HP:0100543"),
  MRUPAV = c("HP:0001618", "HP:0001761", "HP:0002015", "HP:0002460", "HP:0003236", "HP:0003547", "HP:0003555", "HP:0003560", "HP:0003687", "HP:0003691", "HP:0003722", "HP:0003805", "HP:0009027"),
  CPEO = c("HP:0000029", "HP:0000407", "HP:0000505", "HP:0000508", "HP:0000518", "HP:0000565", "HP:0000590", "HP:0000648", "HP:0000651", "HP:0000712", "HP:0000716", "HP:0000786", "HP:0000815", "HP:0000869", "HP:0001260", "HP:0001265", "HP:0001272", "HP:0001284", "HP:0001300", "HP:0001332", "HP:0001350", "HP:0001488", "HP:0001618", "HP:0001634", "HP:0001638", "HP:0001653", "HP:0001761", "HP:0002015", "HP:0002063", "HP:0002066", "HP:0002067", "HP:0002070", "HP:0002078", "HP:0002080", "HP:0002151", "HP:0002322", "HP:0002378", "HP:0002403", "HP:0002460", "HP:0002548", "HP:0002578", "HP:0002650", "HP:0002747", "HP:0002922", "HP:0003200", "HP:0003201", "HP:0003202", "HP:0003236", "HP:0003323", "HP:0003376", "HP:0003390", "HP:0003434", "HP:0003458", "HP:0003487", "HP:0003546", "HP:0003548", "HP:0003551", "HP:0003557", "HP:0003688", "HP:0003689", "HP:0003690", "HP:0003700", "HP:0003701", "HP:0003713", "HP:0003722", "HP:0003737", "HP:0006858", "HP:0006886", "HP:0007641", "HP:0007824", "HP:0008180", "HP:0008209", "HP:0010628", "HP:0030319", "HP:0100295"),
  SCA12 = c("HP:0000317", "HP:0000496", "HP:0000708", "HP:0000716", "HP:0000726", "HP:0000739", "HP:0000746", "HP:0001251", "HP:0001260", "HP:0001272", "HP:0001288", "HP:0001300", "HP:0001310", "HP:0001317", "HP:0001347", "HP:0002059", "HP:0002067", "HP:0002073", "HP:0002075", "HP:0002080", "HP:0002120", "HP:0002174", "HP:0002317", "HP:0002345", "HP:0002346", "HP:0002375", "HP:0002406", "HP:0002530", "HP:0007010", "HP:0007141", "HP:0007256", "HP:0030188", "HP:0100543"),
  HSANVIII = c("HP:0000407", "HP:0000559", "HP:0000633", "HP:0000966", "HP:0001249", "HP:0001265", "HP:0001581", "HP:0002495", "HP:0002579", "HP:0004409", "HP:0007021", "HP:0010829", "HP:0010831", "HP:0012804", "HP:0033748"),
  CJD = c("HP:0000504", "HP:0000505", "HP:0000605", "HP:0000639", "HP:0000712", "HP:0000716", "HP:0000726", "HP:0000736", "HP:0000737", "HP:0000738", "HP:0000739", "HP:0000741", "HP:0000746", "HP:0000751", "HP:0001250", "HP:0001262", "HP:0001269", "HP:0001289", "HP:0001317", "HP:0001324", "HP:0001336", "HP:0001337", "HP:0001350", "HP:0002066", "HP:0002067", "HP:0002072", "HP:0002073", "HP:0002283", "HP:0002312", "HP:0002354", "HP:0002381", "HP:0002401", "HP:0002446", "HP:0002464", "HP:0002529", "HP:0002922", "HP:0003487", "HP:0005327", "HP:0006943", "HP:0007009", "HP:0007017", "HP:0007076", "HP:0007158", "HP:0007183", "HP:0007256", "HP:0007686", "HP:0010542", "HP:0010846", "HP:0011099", "HP:0012332", "HP:0012672", "HP:0025152", "HP:0100256", "HP:0100292", "HP:0100661", "HP:0100785"),
  FAME7 = c("HP:0001250", "HP:0033054"),
  CANVAS = c("HP:0000364", "HP:0000407", "HP:0000639", "HP:0000640", "HP:0000648", "HP:0000750", "HP:0001151", "HP:0001152", "HP:0001251", "HP:0001260", "HP:0001265", "HP:0001272", "HP:0001284", "HP:0001310", "HP:0002066", "HP:0002070", "HP:0002073", "HP:0002075", "HP:0002080", "HP:0002172", "HP:0002403", "HP:0002460", "HP:0002494", "HP:0002495", "HP:0002828", "HP:0003438", "HP:0003447", "HP:0003487", "HP:0006855", "HP:0007108", "HP:0007141", "HP:0007230", "HP:0007328", "HP:0008568", "HP:0009830", "HP:0010545", "HP:0012332", "HP:0012735"),
  OPDM4 = c("HP:0000218", "HP:0000508", "HP:0000544", "HP:0000651", "HP:0001260", "HP:0001284", "HP:0001337", "HP:0001611", "HP:0002015", "HP:0002174", "HP:0002460", "HP:0002465", "HP:0003236", "HP:0003458", "HP:0003557", "HP:0003736", "HP:0003805", "HP:0012548", "HP:0030319"),
  CCD = c("HP:0000162", "HP:0000164", "HP:0000175", "HP:0000189", "HP:0000218", "HP:0000239", "HP:0000242", "HP:0000246", "HP:0000248", "HP:0000256", "HP:0000272", "HP:0000303", "HP:0000316", "HP:0000337", "HP:0000340", "HP:0000347", "HP:0000364", "HP:0000365", "HP:0000389", "HP:0000431", "HP:0000670", "HP:0000680", "HP:0000682", "HP:0000684", "HP:0000696", "HP:0000772", "HP:0000773", "HP:0000774", "HP:0000882", "HP:0000891", "HP:0000894", "HP:0000939", "HP:0001156", "HP:0001172", "HP:0001182", "HP:0001810", "HP:0002007", "HP:0002098", "HP:0002205", "HP:0002643", "HP:0002644", "HP:0002645", "HP:0002650", "HP:0002652", "HP:0002659", "HP:0002684", "HP:0002688", "HP:0002689", "HP:0002700", "HP:0002705", "HP:0002738", "HP:0002757", "HP:0002808", "HP:0002812", "HP:0002827", "HP:0002857", "HP:0002866", "HP:0003183", "HP:0003298", "HP:0003302", "HP:0003304", "HP:0003396", "HP:0004209", "HP:0004220", "HP:0004322", "HP:0004331", "HP:0004474", "HP:0005107", "HP:0005259", "HP:0005280", "HP:0005916", "HP:0005930", "HP:0006040", "HP:0006297", "HP:0006660", "HP:0008391", "HP:0008788", "HP:0008821", "HP:0008848", "HP:0009577", "HP:0010230", "HP:0010535", "HP:0010669", "HP:0010751", "HP:0010807", "HP:0011001", "HP:0011069", "HP:0011219", "HP:0011800", "HP:0100864", "HP:0200021"),
  FAME1 = c("HP:0001249", "HP:0001312", "HP:0001326", "HP:0001337", "HP:0001340", "HP:0001351", "HP:0002069", "HP:0002123", "HP:0010852"),
  XLID_PHPX = c("HP:0000821", "HP:0000824", "HP:0000871", "HP:0001256", "HP:0001263", "HP:0004322"),
  FAME2 = c("HP:0000643", "HP:0000726", "HP:0001249", "HP:0001251", "HP:0001312", "HP:0001326", "HP:0001336", "HP:0001337", "HP:0001340", "HP:0001351", "HP:0002069", "HP:0002076", "HP:0010852", "HP:0100543"),
  XDP = c("HP:0000643", "HP:0001288", "HP:0001304", "HP:0001336", "HP:0001337", "HP:0002067", "HP:0002072", "HP:0002172", "HP:0002322", "HP:0002359", "HP:0002362", "HP:0002378", "HP:0002451", "HP:0002548", "HP:0004373", "HP:0006511", "HP:0007158", "HP:0010808", "HP:0011951", "HP:0031162"),
  SCA17 = c("HP:0000020", "HP:0000473", "HP:0000640", "HP:0000643", "HP:0000708", "HP:0000716", "HP:0000718", "HP:0000726", "HP:0000727", "HP:0000738", "HP:0000743", "HP:0000757", "HP:0001250", "HP:0001251", "HP:0001257", "HP:0001260", "HP:0001268", "HP:0001272", "HP:0001288", "HP:0001289", "HP:0001300", "HP:0001310", "HP:0001332", "HP:0001336", "HP:0002015", "HP:0002063", "HP:0002066", "HP:0002067", "HP:0002070", "HP:0002072", "HP:0002080", "HP:0002136", "HP:0002171", "HP:0002186", "HP:0002300", "HP:0002356", "HP:0002403", "HP:0002506", "HP:0002529", "HP:0004305", "HP:0007058", "HP:0007256", "HP:0007366", "HP:0007668", "HP:0011999", "HP:0012082"),
  TOF = c("HP:0000028", "HP:0000233", "HP:0000268", "HP:0000337", "HP:0000520", "HP:0001156", "HP:0001511", "HP:0001636", "HP:0004209", "HP:0004467", "HP:0005105", "HP:0009891"),
  FECD3 = c("HP:0000505", "HP:0007957", "HP:0012038", "HP:0012040"),
  FAME6 = "HP:0033054",
  HMNR7 = c("HP:0001252", "HP:0001265", "HP:0001308", "HP:0001371", "HP:0001761", "HP:0001762", "HP:0002359", "HP:0003326", "HP:0003401", "HP:0003458", "HP:0003691", "HP:0007002", "HP:0007210", "HP:0008959", "HP:0008994", "HP:0008997", "HP:0009027", "HP:0009053", "HP:0010830"),
  DBQD2_BSS = c("HP:0000175", "HP:0000193", "HP:0000252", "HP:0000280", "HP:0000286", "HP:0000311", "HP:0000343", "HP:0000470", "HP:0000520", "HP:0000592", "HP:0000664", "HP:0000668", "HP:0000678", "HP:0000750", "HP:0000767", "HP:0000768", "HP:0000774", "HP:0000894", "HP:0000926", "HP:0000954", "HP:0001249", "HP:0001252", "HP:0001263", "HP:0001382", "HP:0001511", "HP:0001763", "HP:0001769", "HP:0001840", "HP:0001863", "HP:0001956", "HP:0002643", "HP:0002656", "HP:0002673", "HP:0002827", "HP:0002938", "HP:0002970", "HP:0003016", "HP:0003026", "HP:0003048", "HP:0003180", "HP:0003510", "HP:0004233", "HP:0004482", "HP:0004976", "HP:0005280", "HP:0008897", "HP:0009803", "HP:0010049", "HP:0011304", "HP:0012368", "HP:0012725", "HP:0031936", "HP:0033102"),
  FAME4 = c("HP:0001249", "HP:0001250", "HP:0001336", "HP:0001337", "HP:0001340", "HP:0001351", "HP:0002069", "HP:0002392"),
  SCA4 = c("HP:0000763", "HP:0001251", "HP:0001260", "HP:0001265", "HP:0001272", "HP:0001284", "HP:0001288", "HP:0002073", "HP:0002333", "HP:0002406", "HP:0002495", "HP:0002936", "HP:0003390", "HP:0003438", "HP:0003487", "HP:0007002", "HP:0007772", "HP:0009830", "HP:0010830", "HP:0010831"),
  HPE5 = c("HP:0000202", "HP:0000218", "HP:0000238", "HP:0000243", "HP:0000252", "HP:0000316", "HP:0000337", "HP:0000340", "HP:0000341", "HP:0000348", "HP:0000400", "HP:0000463", "HP:0000520", "HP:0000582", "HP:0000601", "HP:0000664", "HP:0000863", "HP:0001249", "HP:0001250", "HP:0001263", "HP:0001272", "HP:0001347", "HP:0001360", "HP:0002002", "HP:0002119", "HP:0002507", "HP:0005280", "HP:0006870", "HP:0006956", "HP:0006988", "HP:0012385", "HP:0025670", "HP:0032549", "HP:0100702"),
  VACTERLX = c("HP:0000068", "HP:0000105", "HP:0000126", "HP:0000238", "HP:0000925", "HP:0000960", "HP:0001161", "HP:0001561", "HP:0001651", "HP:0001669", "HP:0002023", "HP:0002032", "HP:0002089", "HP:0002575", "HP:0003468", "HP:0003974", "HP:0005301", "HP:0005792", "HP:0006695", "HP:0008750", "HP:0009623", "HP:0010963", "HP:0030680", "HP:0031853")
  )
  # Removed "HP:6000414" (duplicate uterine cervix) from HFGI, HFGII, HFGIII because not included in data(hpo) ontology_index, but is in hpo.jax.org
  # Removed "HP:0020068" (Intranuclear inclusion bodies) from NIID because not included in data(hpo) ontology_index, but is in hpo.jax.org
  # Removed "HP:6000872" (Aplasia/Hypoplasia of the calvaria) and "HP:6000875" (Wide sacroiliac joint) from CCD because not included in data(hpo) ontology_index, but is in hpo.jax.org
  # Removed "HP:6000816" (Prominent lesser trochanter) from DBQD2_BSS because not included in data(hpo) ontology_index, but is in hpo.jax.org
  sim_mat <- get_sim_grid(ontology = hpo, term_sets = term_set
  )
  sim_mat
  dist_mat <- max(sim_mat) - sim_mat
  plot(hclust(as.dist(dist_mat))
  )
  hclust_obj <- hclust(as.dist(dist_mat)
  )
  pheatmap(sim_mat,
           cluster_rows = hclust_obj,
           cluster_cols = hclust_obj,
           show_rownames = TRUE,
           show_colnames = TRUE,
           fontsize = 6.5
  )
```








## Visualizing STR motifs (incomplete)

Visualizations of pathogenic_motif_reference_orientation using `ggplot2` functions

```{r}
#| label: visualize "pathogenic_motif_reference_orientation" motifs


# Load packages
library(jsonlite)
library(tidyverse)

# Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Select relevant data for heat map visualization
hpo_hm_data <- loci |> select(hpo_terms, disease_id, pathogenic_motif_reference_orientation)
view(hpo_hm_data)

# Select the most frequent motifs in "pathogenic_motif_reference_orientation" data column (based on frequency across disease_id)
most_frequent_motifs <- loci |> select(pathogenic_motif_reference_orientation, disease_id) |> count(pathogenic_motif_reference_orientation, sort = T)
view(most_frequent_motifs)
```