---
title: "STRchive Loci: HPO Terms Data Analysis"
format: html
editor: visual
---

## Importing STRchive-loci.json Data and Accessing HPO Terms

Going to read STRchive data from GitHub into R using the `fromJSON()` function from the `jsonlite` package. Then will view the data using the `View()` function and access the nested HPO terms from the data using the `$` command.

```{r}
#| label: read-jsonlite

# Load packages
library(jsonlite)

# Read data from url into an object called "loci"
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# View the data
View(loci)

# Select HPO terms
loci$hpo_terms
```

## Example: Creating a Plot with STR Motif Length Data

Create a bar plot using the `table()` and `barplot()` functions.

```{r}
#| label: create bar plot

# Load packages
library(jsonlite)

#Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Create barplot
counts <- table(loci$motif_len)
barplot(counts, main = "STRchive loci by motif length",
        xlab = "Motif length (bp)",
        ylab = "Number of loci",
        col = "pink")
```

## Consolidating the "loci" Data Frame into HPO Terms

Going to remove columns from "loci" data using the `select()` function from the `dplyr` package so that only disease_id and hpo_terms are included.

```{r}
#| label: consolidate "loci" data into hpo terms

# Load packages
library(dplyr)
library(jsonlite)

# Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Remove columns from loci data table
hpo_terms <- loci %>% select(c(disease_id, hpo_terms))
View(hpo_terms)
```

## Unnesting and sorting "hpo_terms" from hpo_terms data frame

Going to unnest the hpo_terms from the disease_id rows in the hpo_terms data frame using the `unnest_longer()` function from the `tidyverse` package so that the individual phenotypes for each disease are accessible. Also going to use `relocate` to rearrange data frame with hpo_terms at the front. Finally, going to create a separate data frame of phenotypes sorted by frequency using the `count()` function.

```{r}
#| label: unnest "hpo_terms" values from "disease_id" rows

# Load packages
library(jsonlite)
library(tidyverse)

# Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Remove columns from loci data table and create new hpo_terms data frame
hpo_terms <- loci %>% select(c(disease_id, hpo_terms))
View(hpo_terms)

# Unnest hpo_terms, relocate hpo_terms, and create new unnested_hpo_terms data frame
unnested_hpo_terms <- hpo_terms |> unnest_longer(hpo_terms) |> relocate(hpo_terms)
view(unnested_hpo_terms)

# Sort hpo_terms by frequency
sorted_hpo_terms <- unnested_hpo_terms |> count(hpo_terms, sort = T)
view(sorted_hpo_terms)
```

## Visualizing hpo_terms

Different visualizations of hpo_terms using different `ggplot` functions form the `gglot2` package.

```{r}
#| label: visualize "hpo_terms" as different plots

# Load packages
library(jsonlite)
library(tidyverse)
library(ontologyIndex)

# Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Consolidate and unnest hpo_terms
hpo_terms <- loci |> select(c(disease_id, hpo_terms))
unnested_hpo_terms <- hpo_terms |> unnest_longer(hpo_terms) |> relocate(hpo_terms)

# Visualize all hpo_terms vs. frequency as a bar plot
unnested_hpo_terms |> ggplot(aes(x = hpo_terms)) + geom_bar(colour = "lightgreen") + labs(title = "HPO Frequency", subtitle = "Frequency of phenotypes across different STR diseases", x = "HPO Term", y = "Frequency")

# Visualize hpo_terms with a frequency of 13+ across diseases
most_frequent_hpo <- unnested_hpo_terms |> count(hpo_terms, sort = T) |> filter(n >= 13)
ggplot(most_frequent_hpo, aes(x = n, y = fct_infreq(hpo_terms, (n)) |> fct_rev())) + geom_col(fill = "lightblue") + labs(title = "Most Frequent HPOs", subtitle = "Phenotypes with a frequency of 13+ across STR diseases", x = "Frequency", y = "HPO Term")

# Visualize hpo_terms with a frequency of 13+ across diseases and with "Ataxia" types consolidated (Based on "children", or `get_descendants()` using `ontologyIndex` package with `data(hpo)`
hpo_frequency <- unnested_hpo_terms |> mutate(
  hpo_terms = fct_recode(hpo_terms,
  "HP:0001251 Ataxia" = "HP:0001251 Ataxia",
  "HP:0001251 Ataxia" = "HP:0001310 Dysmetria",
  "HP:0001251 Ataxia" = "HP:0002066 Gait ataxia",
  "HP:0001251 Ataxia" = "HP:0002070 Limb ataxia",
  "HP:0001251 Ataxia" = "HP:0002073 Progressive cerebellar ataxia",
  "HP:0001251 Ataxia" = "HP:0002075 Dysdiadochokinesis",
  "HP:0001251 Ataxia" = "HP:0002078 Truncal ataxia",
  "HP:0001251 Ataxia" = "HP:0010867 Dyssynergia",
  "HP:0001251 Ataxia" = "HP:0002070 Limb Ataxia",
  )) |> count(hpo_terms, sort = T) |> filter (n >= 13)
ggplot(hpo_frequency, aes(x = n, y = fct_infreq(hpo_terms, (n)) |> fct_rev())) + geom_col(fill = "lightblue") + labs(title = "Most Frequent HPOs", subtitle = "Phenotypes with a frequency of 13+ across STR diseases", x = "Frequency", y = "HPO Term")

## FINAL HPO FREQUENCY PLOT
# Made using `fct_collapse()` function and manual matching with "children" from `ontologyIndex` data(hpo)
hpo_frequency <- unnested_hpo_terms |> mutate(
  hpo_terms = fct_collapse(hpo_terms,
  "HP:0001251 Ataxia" = c("HP:0001251 Ataxia", "HP:0001310 Dysmetria", "HP:0002066 Gait ataxia", "HP:0002070 Limb ataxia", "HP:0002073 Progressive cerebellar ataxia", "HP:0002075 Dysdiadochokinesis", "HP:0002078 Truncal ataxia", "HP:0010867 Dyssynergia", "HP:0002070 Limb Ataxia"),
  "HP:0001288 Gait disturbance" = c("HP:0001288 Gait disturbance", "HP:0002136 Broad-based gait", "HP:0002141 Gait imbalance", "HP:0002317 Unsteady gait", "HP:0002362 Shuffling gait", "HP:0002515 Waddling gait", "HP:0002527 Falls", "HP:0002540 Inability to walk", "HP:0003376 Steppage gait", "HP:0030051 Tip-toe gait", "HP:0031629 Impaired tandem gait"),
  "HP:0001260 Dysarthria" = c("HP:0001260 Dysarthria", "HP:0002464 Spastic dysarthria", "HP:0007024 Pseudobulbar paralysis", "HP:0008376 Nasal dysarthria"),
  "HP:0002015 Dysphagia" = c("HP:0002015 Dysphagia", "HP:0007024 Pseudobulbar paralysis", "HP:0031162 Impaired oropharyngeal swallow response", "HP:0200136 Oral-pharyngeal dysphagia"),
  "HP:0000639 Nystagmus" = c("HP:0000639 Nystagmus", "HP:0000640 Gaze-evoked nystagmus", "HP:0000666 Horizontal nystagmus", "HP:0010542 Vestibular nystagmus", "HP:0010544 Vertical nystagmus"),
  "HP:0001272 Cerebellar atrophy" = c("HP:0001272 Cerebellar atrophy", "HP:0006855 Cerebellar vermis atrophy", "HP:0007263 Spinocerebellar atrophy", "HP:0012082 Cerebellar Purkinje layer atrophy", "HP:0100275 Diffuse cerebellar atrophy"),
  "HP:0001249 Intellectual disability" = c("HP:0001249 Intellectual disability", "HP:0001256 Mild intellectual disability", "HP:0002342 Moderate intellectual disability", "HP:0006889 Borderline intellectual disability", "HP:0010864 Severe intellectual disability"),
  "HP:0001337 Tremor" = c("HP:0001337 Tremor", "HP:0002322 Resting tremor", "HP:0002345 Action tremor", "HP:0030188 Tremor by anatomical site"),
  "HP:0001250 Seizure" = c("HP:0001250 Seizure", "HP:0002069 Bilateral tonic-clonic seizure", "HP:0002133 Status epilepticus"),
  "HP:0001347 Hyperreflexia" = c("HP:0001347 Hyperreflexia", "HP:0001348 Brisk reflexes", "HP:0002169 Clonus", "HP:0006801 Hyperactive deep tendon reflexes"),
  "HP:0000726 Dementia" = c("HP:0000726 Dementia", "HP:0000727 Frontal lobe dementia", "HP:0002145 Frontotemporal dementia", "HP:0007123 Subcortical dementia"),
  "HP:0001265 Hyporeflexia" = ("HP:0001265 Hyporeflexia"),
  "HP:0002067 Bradykinesia" = ("HP:0002067 Bradykinesia"),
  "HP:0003487 Babinski sign" = ("HP:0003487 Babinski sign")
  )) |> count(hpo_terms, sort = T) |> filter (n >= 13)

ggplot(hpo_frequency, aes(x = n, y = fct_infreq(hpo_terms, (n)) |> fct_rev())) + geom_col(fill = "lightblue") + labs(title = "Most Frequent HPOs", subtitle = "Phenotypes with a frequency of 13+ across STR diseases", x = "Frequency", y = "HPO Term")

```

## Heatmapping with ontologyX Package Data

Using `onotologySimilarity` and `ontologyIndex` functions to create similarity matrices and distance clusters to then be visualized with `pheatmap`

```{r}
#| label: Similarity matrices, cluster dendrograms, and heat maps of HPO data


# Load packages and similarity matrix setup
library(pheatmap)
library(ontologyIndex)
library(ontologySimilarity)
data(hpo)
set.seed(1)
information_content <- descendants_IC(hpo)

# Create similarity matrix
  # First generate terms
  term_sets <- list(
  Ataxia = get_descendants(hpo, "HP:0001251"),
  Gait_Ataxia = get_descendants(hpo, "HP:0002066"),
  Gait_Disturbance = get_descendants(hpo, "HP:0001288"),
  Seizure = get_descendants(hpo, "HP:0001250")
  )
  # Next, calculate simialrity matrix
  sim_mat <- get_sim_grid(ontology = hpo, term_sets = term_sets)
  sim_mat
  # Convert to distance matrix (cluster dendrogram), which is necessary to visualize as   heat map 
  dist_mat <- max(sim_mat) - sim_mat
  plot(hclust(as.dist(dist_mat)))
  # Visualize as heatmap
   hclust_obj <- hclust(as.dist(dist_mat))
   pheatmap(sim_mat,
           cluster_rows = hclust_obj,
           cluster_cols = hclust_obj,
           show_rownames = TRUE,
           show_colnames = TRUE)
   
## BIG HEATMAP
  # Manually input uncollapsed hpo_freq terms
  library(ontologyIndex)
  library(ontologySimilarity)
  data(hpo)
  set.seed(50)
  information_content <- descendants_IC(hpo)
# First, similarity matrix of individual terms  
  # Generate terms
  term_sets <- list(
    Dysarthria = "HP:0001260",
    Ataxia = "HP:0001251",
    Dysphagia = "HP:0002015",
    Cerebellar_Atrophy = "HP:0001272",
    Gait_Disturbance = "HP:0001288",
    Nystagmus = "HP:0000639",
    Intellectual_Disability = "HP:0001249",
    Tremor = "HP:0001337",
    Dementia = "HP:0000726",
    Seizure = "HP:0001250",
    Hyporeflexia = "HP:0001265",
    Hyperreflexia = "HP:0001347",
    Gait_Ataxia = "HP:0002066",
    Limb_Ataxia = "HP:0002070",
    Bradykinesia = "HP:0002067",
    Babinski_Sign = "HP:0003487",
    Progressive_Cerebellar_Ataxia = "HP:0002073",
    Depression = "HP:0000716",
    Hypotonia = "HP:0001252",
    Global_Developmental_Delay = "HP:0001263",
    Dysmetria = "HP:0001310",
    Muscle_Weakness = "HP:0001324",
    Dystonia = "HP:0001332",
    Myoclonus = "HP:0001336",
    Ptosis = "HP:0000508",
    Rigidity = "HP:0002063",
    Distal_Muscle_Weakness = "HP:0002460",
    Skeletal_Muscle_Atrophy = "HP:0003202",
    Spasticity = "HP:0001257",
    Parkinsonism = "HP:0001300"
  )
  # Similarity matrix
  sim_mat <- get_sim_grid(ontology = hpo, term_sets = term_sets)
  sim_mat
  # Cluster dendrogram
  dist_mat <- max(sim_mat) - sim_mat
  plot(hclust(as.dist(dist_mat)))
  # Heat map
  hclust_obj <- hclust(as.dist(dist_mat))
  pheatmap(sim_mat,
           cluster_rows = hclust_obj,
           cluster_cols = hclust_obj,
           show_rownames = TRUE,
           show_colnames = TRUE,
           fontsize = 6.5
           )
```








## Visualizing STR motifs (incomplete)

Visualizations of pathogenic_motif_reference_orientation using `ggplot2` functions

```{r}
#| label: visualize "pathogenic_motif_reference_orientation" motifs


# Load packages
library(jsonlite)
library(tidyverse)

# Read data from url into "loci" object
url <- "https://raw.githubusercontent.com/dashnowlab/STRchive/main/data/STRchive-loci.json"
loci <- fromJSON(url)

# Select relevant data for heat map visualization
hpo_hm_data <- loci |> select(hpo_terms, disease_id, pathogenic_motif_reference_orientation)
view(hpo_hm_data)

# Select the most frequent motifs in "pathogenic_motif_reference_orientation" data column (based on frequency across disease_id)
most_frequent_motifs <- loci |> select(pathogenic_motif_reference_orientation, disease_id) |> count(pathogenic_motif_reference_orientation, sort = T)
view(most_frequent_motifs)
```